\documentclass{article}[12pt]
\input{format.tex}
\title{Modeling Tonal Effects of Vowel Deletion as Logical
Transductions}
\date{}
\begin{document}
\maketitle
\vspace{-2em}
\textbf{Introduction} This paper presents a formal proposal to 
model the tonal processes resulting from vowel deletion. We argue
that such processes can be reduced to two operations over 
association relations: \textbf{deassociation} and 
\textbf{reassociation} which is consistent with the core idea 
of autosegmental phonology \citep{goldsmith1976}. 
This lays a foundation for learning more complex tonal processes 
over multitiered structures.
We focus on three languages that represent major tonal 
patterns triggered by vowel deletion in hiatus: Engenni, 
Estako, and Ogoja Yala \citep{kao2017phonological}. In all 
three languages, the first vowel in the hiatus is deleted, 
but the associated tonal outcomes differ. We propose all 
three processes can be captured by a set of transduction 
relations mapping inputs to outputs and differ only in 
the language-specific association line modifications.

\ea 
\textbf{Engenni (tone deletion):} 
Deleting the vowel also removes its associated tone.
\vspace{-1em}
\begin{multicols}{2}
\quad\ L\quad H\qquad\qquad H\\
 /\` ɔkò \# édèì/ $\rightarrow$ [\` ɔkédèì] \\
 \textit{man \ \ canoe} $\rightarrow$ \textit{``a man’s canoe"} 
 
 \quad\ \ H\quad L\qquad\qquad\ \ \ L\\
/fùnú \# èdhí/ $\rightarrow$ [fùnèdhí] \\
\textit{climb \ \ palm tree} $\rightarrow$ \textit{``to climb a palm tree"}
\end{multicols}
\z \vspace{-1em}
\ea \textbf{Estako (contour formation):} 
Deleted vowel’s tone reassociates to the surviving vowel to form a contour.
\vspace{-1em}
\begin{multicols}{2}
\quad\ \ L \ \ \ H\qquad\qquad LH\\
/ówà \# ówà/ \ $\rightarrow$ [ówǒwà] \\
\textit{house \ \ house} $\rightarrow$ \textit{``every house"}

\qquad\ H \ \ \ L\qquad\qquad HL\\
/\`ɔ \# dé \ \# àkpà/ $\rightarrow$ [\`ɔdâkpà] \\
\textit{he \ \ buy \ \ cup}\quad$\rightarrow$ \textit{``he buys a cup"}
\end{multicols}
\z\vspace{-1em}
\ea \textbf{Ogoja Yala (higher-pitch preservation)}
H tone is preserved. 
If no H is present (i.e., only L and M), then M is preferred.
The preserved tone either replaces the tone of the surviving vowel 
(if surviving vowel has a lower pitched tone), or stay unchanged.
\vspace{-1em}
\begin{multicols}{3}
\quad L \; \  H \qquad \quad \ H\\
/dè \# áchí/ $\rightarrow$ [dáchí]\\
\textit{give \ grass} $\rightarrow$ \textit{``to give grass"}

\quad H\quad L\qquad\ \ \ \quad H\\
/má \# \` ɔchí/ $\rightarrow$ [m\' ɔchí]\\
\vspace{1em}\textit{see \quad tree} $\rightarrow$ \textit{``to see tree"}

\ \ M \ \ L \qquad \quad \ M\\
/b\=i \# \` ɔchí/ $\rightarrow$ [b\=ɔchí]\\
\textit{carry  stick} $\rightarrow$ \textit{``to carry a stick"}
\end{multicols}
\z\vspace{-1em}

\paragraph{Logical Transduction of Tonal Processes}

Using model theory
\citep{strother2019diss,yolyan2025logical,jardine2017local,oakden2020notation,jardineDanis2021Qtheory,ChandleeJardine2021BMRS},
we define a phonological representation as relational structures of
the form $\langle D, \mathcal{R} \rangle$, where $D$ is a finite
domain (a finite set of positions), and $\mathcal{R}$ is a set of
relations (also known as the model \textit{signature}).  An
autosegmental representation (AR) is defined over a signature
consisting of unary predicates, which indicate tonal or segmental
elements, such as $H, L, M,$ and $C, V$, and binary relations that encode
structural connections. In particular, we assume two binary relations:
$\lhd$ and $\alpha$. The relation $\lhd$ encodes the successor
relation: $\lhd(x,y)$ indicates that $x$ immediately proceeds $y$. For
example, vowel hiatus can be defined as
$\textsf{hiatus}(x,y) := V(x) \land V(y) \land \lhd(x,y)$, which
refers to a phonological context where $x$ and $y$ are both vowels
and $x$ precedes $y$. In the three language examples considered here,
the first vowel $x$ is consistently deleted (denoted
\textsf{delete\_V}), while the second vowel $y$ survives (denoted
\textsf{survive\_V}). In the case of Ogoja Yala, it will also be
important to pick out whether the first of two successive tones bears
a higher pitch, for which we use the predicate
$\textsf{higher}(x,y) := \lhd(x,y) \land \big((H(x) \land (M(y) \lor L(y))) \lor (M(x)
\land L(y))\big)$. 

\input{ARmodels.tex} The relation $\alpha(x,y)$ encodes across-tier 
association, indicating 
that tone $x$ links to segment $y$. We require that the association relation be symmetric, i.e.,
$\alpha(x, y) \iff \alpha(y, x)$.
For convenience, we adopt the convention that the first 
variable $x$ refers to the tone, and $y$ refers to the segment.
These predicates, defined 
in model-theoretic terms, mathematically capture the autosegmental relation \citep{goldsmith1976,coleman1991no}.

Building on this view, we define a phonological transformation 
from an underlying representation (UR) to a surface 
representation (SR) is modeled as a relation between input
and output structures.
The output structure is obtained by evaluation of formulas with 
respect to the input structure.
For example, for any predicate $R(x)$ in the input signature,
a faithful output predicate $R'$ is defined as $R'(x) := R(x)$,
meaning that an element $x$ in the output domain has property $R$
iff it has property $R$ in the input. 

\textbf{Our proposal} is that the processes in question preserve all relations from input to output except 
for $\alpha$, which is subject to language-specific 
\textit{deassociation}, and 
\textit{reassociation}:
\[
\alpha'(x,y) := (\alpha(x,y)\land \neg \textsf{deassoc}(x,y)) \lor \textsf{reassoc}(x,y)
\]
Table~\ref{tab.transduction} provides the transductions.
The unassociated units in the domain are not realized at the surface,
since the licensing function $\phi_{\textsf{license}}$ restricts
realization to fully associated units.
Unary relations are preserved faithfully.
The successor relation $\lhd$ is faithful modulo unlicensed units,
handled by order preservation \citep{bhaskar+2023}.

 \input{tab.transduction.tex} 

The output association relation $\alpha'$ is defined
as follows: 
an association line holds in the output if (i) it is present 
in the input and not removed by \textsf{deassoc}, or (ii) it is 
introduced by \textsf{reassoc}. Concretely, \textsf{deassoc} 
removes the association between the deleted vowel and its tone, or as in the case of 
Ogaja Yala, between the surviving vowel
and its associated L tone, if 
there is one.
The \textsf{reassoc} operation then determines 
the outcome: it may be inactive (Engenni), reattach the tone to 
the surviving vowel (Estako), or rearrange the association depending
on the relative pitch of two tones (Ogoja Yala).

\input{tab.all.tex} \textbf{(a) Tone deletion in Engenni} is derived
by a single operation, \textsf{deassoc}, which deassociates the
deleted vowel and its tone (indicated by the crossed-out line in
Table~\ref{tab2}).  This leaves both $T_1$ and $V_1$
unassociated. Since \textsf{reassoc} is inactive, no new association
is formed. Consequently, $T_1$ and $V_1$ remain unassociated,
therefore are not licensed by $\phi_{\textsf{license}}$ (indicated in
gray in Table~\ref{tab2}). Thus, only $T_2$ and $V_2$ surface in the
output.

\textbf{(b) Contour formation in Estako} is derived by the same 
\textsf{deassoc} operation. The \textsf{reassoc} operation then 
links the tone of the deleted vowel ($T_1$) to the surviving 
vowel ($V_2$, as indicated by the dotted line in Table \ref{tab2}). 
As a result, $V_2$ becomes associated with both $T_1$ and $T_2$, 
which forms a contour tone. At this point, only $V_1$ is unassociated 
and therefore not licensed. The surface form thus bears a contour 
tone on the surviving vowel.

\textbf{(c) Higher-pitch preservation in Ogoja Yala} preserves the 
higher-pitched tone by associating it with the surviving vowel. The 
\textsf{deassoc} operation first breaks the association between the 
deleted vowel and its tone, regardless of pitch. It then evaluates the 
tone on the surviving vowel: (i) if its pitch is lower than the 
preceding tone, this association is also broken, and \textsf{reassoc} 
links the surviving vowel to the higher-pitched tone (the preceding 
one); (ii) if its pitch is higher, the surviving vowel keeps its 
original association, and no further \textsf{reassoc} applies (as in 
Engenni). Similarly, any units not associated are not licensed.

\textbf{Conclusion} This paper has argued that vowel-deletion–induced
tonal processes can be uniformly captured as logical transductions
over ARs. The observed typological variation follows from modifying a
single relation, $\alpha$, through the interaction of \textsf{deassoc}
and \textsf{reassoc}.  A limitation of the current analysis is that
the transduction relies on quantification to capture across-tier
dependencies. An important direction for future work is to determine
whether these processes can be simplified to a quantifier-free (QF)
process \citep{strother2018imdlawn, oakden2020notation}.  This work
also extends naturally to learning over ARs. Recent work by
\citet{yolyan2025logical} demonstrates that logical transductions for
segmental phonology can be learned by adapting ideas from
\citet{chandleeBufia2019}, and extending that work to autosegmental
phonology is an important next step.

\section{Learning association changes using logical transductions}
As mentioned above, the three types under discussion can be captured through a single association relation
\(\alpha\) together with two sub-operations, \textsf{deassoc} and
\textsf{reassoc}, modeled as logical transductions. These
transductions are grounded in model-theoretic representations of
phonological structures, where logical formulas define the mapping
from input to output representations.

As \citep{yolyan2025logical} argues, a major advantage of this approach
is that it allows much greater flexibility in phonological
representation for learning compared with traditional finite-state
transducers, which primarily operate over strings. She further shows
that model-theoretic representations facilitate learning by allowing
the learner to search through a partially ordered space of possible
structures, rather than through a purely linear string space.
In this paper, we show how this learning framework can be extended to
learn tonal processes in which the primary changes involve
modifications of association lines rather than changes to segments or
features themselves.

It should be noted that although our discussion is rooted in tonal 
phonology, the same learning approach can in principle be applied 
to any phonological representation, provided that the input and output 
structural relations are clearly defined model-theoretically. 

\section{Case Study: Estako Accusative Compound}
We illustrate the learning process using the Estako contour 
formation pattern as a case study. Let us assume a small 
dataset consisting of input-output pairs, specifically, 
a accusative noun phgrases.

In Estako, the noun in accusative form always ends with High tone.


\begin{table}[h!]
\centering
\begin{tabular}{llll}
\toprule
\textbf{Compound} & \textbf{Underlying Form} & \textbf{Surface Form} & \textbf{Gloss} \\
\midrule
a. /ówà/ && [ówà] & `house-singular=stem' \\
b. /ówá/ && [ówá] & `house-ACC' \\
% /áwé/ + /mhè/ & /áwémhè/ & [áwémhè] & `my water' \\
c. /ówá/ + /mhè/ & /ówámhè/ & [ówámhè] & `my house' \\
d. /ówá/ + /è\texttheta à/ & /ówáè\texttheta à/ & [ówê\texttheta à] & `father's house' \\
e. /ówà/ + /ówà/ & /ówàówà/ & [ówǒwà] & `every house' \\
\bottomrule
\end{tabular}
\caption{Possessive tonal alternations}
\end{table}

As one can see that a b c all surface faithfully to the underlying.
D and e, when compounding together, resutls in a vowel hiatus,
and will delete the first vowel, and do a contour. Instead of
drawing the AR graph, we represent this changes in matrices.
And abstract it away from the segemntal information, just use
C and V, H and L.

$
\begin{array}{c|ccc}
& H & L \\
\hline
V & 1 & 0 \\
C & 0 & 0 \\
V & \colorbox{pink}{1} & 0\\
V & \colorbox{pink}{0} & 1\\
C & 0 & 0 \\
V & 0 & 1 \\
\end{array}
$$
\begin{array}{c|ccc}
& H & L \\
\hline
V & 1 & 0 \\
C & 0 & 0 \\
V & \colorbox{pink}{0} & 0\\
V & \colorbox{pink}{1} & 1\\
C & 0 & 0 \\
V & 0 & 1 \\
\end{array}
$\hfill $
\begin{array}{c|cccc}
& H & L & H & L \\
\hline
V & 1 & 0 & 0 & 0\\
C & 0 & 0 & 0 & 0\\
V & 0 & \colorbox{pink}{1}  & 0 & 0\\
V & 0 & \colorbox{teal}{0} & 1 & 0\\
C & 0 & 0 & 0 & 0\\
V & 0 & 0 & 0 & 1\\
\end{array}
$
\quad $
\begin{array}{c|cccc}
& H & L & H & L \\
\hline
V & 1 & 0 & 0 & 0\\
C & 0 & 0 & 0 & 0\\
V & 0 & \colorbox{pink}{0}  & 0 & 0\\
V & 0 & \colorbox{teal}{1} & 1 & 0\\
C & 0 & 0 & 0 & 0\\
V & 0 & 0 & 0 & 1\\
\end{array}
$

There are two changes from input to output. First, the association
between the first vowel in the hiatus and its tone is removed. This is
a \textsc{deassociation} operation, indicated by the value in the cell
changing from \(1\) to \(0\) (shown in pink). Second, the association
between the second vowel and its tone is added. This is a
\textsc{reassociation} operation, indicated by the change from \(0\) to
\(1\) (shown in green). All other associations remain unchanged.

From segmental phonology, we are familiar with rewrite rules such as

\[
A \rightarrow B \,/\, C \_ D
\]

where \(C \_ D\) represents the contextual environment that triggers
the change from \(A\) to \(B\). In autosegmental phonology, such
processes are not always easy to express linearly, since the relevant
structure involves relations rather than simple strings. However, the
learning question remains the same: what environment triggers the
change from \(1\) to \(0\), and what environment triggers the change
from \(0\) to \(1\)?

With only a single UR--SR pair, there are infinitely many possible
hypotheses. For example, one might hypothesize that all \(H\) tones
delete their associations, or that all vowels do. However, once more
data points are observed, counterexamples eliminate incorrect
generalizations.

The learning task, therefore, is to determine the surrounding
environment of a particular cell value—either \(1\) or \(0\)—that
triggers the change.

\[
\begin{array}{ccc}
? & ? & ? \\
? & \colorbox{pink}{0} & ? \\
? & ? & ?
\end{array}
\quad \rightarrow \quad
\begin{array}{ccc}
? & ? & ? \\
? & \colorbox{pink}{1} & ? \\
? & ? & ?
\end{array}
\]

% \begin{tikzpicture}[scale=0.7, every node/.style={inner sep=2pt}]
%   \node [fill=gray!20](v1) at (0,0) {V\textsubscript{1}};
%   \node (v2) at (1.2,0) {V\textsubscript{2}};
%   \node (t1) at (0,1.2) {T\textsubscript{1}};
%   \node (t2) at (1.2,1.2) {T\textsubscript{2}};
%   \draw (v1)--(t1) node[midway] {=};
%   \draw (v2)--(t2);
%   \draw[dotted, thick] (v2)--(t1);
% \end{tikzpicture} 
\subsection{Learning process}

We begin with an idealized assumption: the tonal process involves no
change to the segmental string, and the only modifications between the
underlying representation (UR) and the surface representation (SR) are
changes in association lines, namely \textsc{deassociation} and
\textsc{reassociation}. Although many phonological processes also
involve segmental deletion, insertion, or feature changes, we restrict
the current discussion to association-line modifications in order to
isolate the learning problem.

The first step is to identify exactly what structural modification has
taken place between the UR and the SR. Since the segmental and tonal
tiers remain constant, the comparison is made directly over the
association relation matrix.

For each cell in the association matrix:

\[
\text{if } \mathrm{UR}_{ij} = 1 \text{ and } \mathrm{SR}_{ij} = 0,
\quad \Rightarrow \quad \textsc{deassociation}
\]

\[
\text{if } \mathrm{UR}_{ij} = 0 \text{ and } \mathrm{SR}_{ij} = 1,
\quad \Rightarrow \quad \textsc{reassociation}
\]

That is, if an association present in the UR disappears in the SR, the
operation is classified as deassociation; if a new association appears
in the SR that was absent in the UR, the operation is classified as
reassociation.

In Estoko, both operations are attested: one association line is
removed while another is added. Therefore, the learner must infer both
the deassociation environment and the reassociation environment
separately. (maybe not maybe they can start together)

\paragraph{Hypothesis space for deassociation.}

Once a deassociation site has been identified, the learner constructs
candidate structural descriptions for the environment in which a
\(1\)-valued association changes to \(0\). The search begins with the
smallest possible factors and gradually adds more structure. Each
candidate factor corresponds to a hypothesis about what triggers
deassociation.

The most general hypothesis is that every existing association is
removed:

\[
\text{Hypothesis 1: } 1 \rightarrow 0
\]

This hypothesis is too general, since there are counterexamples in
which other \(1\)-valued associations remain unchanged, such as
\((H,V_1)\), \((L,V_3)\), and \((L,V_4)\).

A slightly more specific hypothesis is that deassociation is triggered
by a single element from either tier: either a particular tone or a particular segment that the
changing association line is connecting with.
This gives the following one-factor hypotheses:

\[
\begin{array}{c}
% \text{one-factor with only an } H \text{ column}
% \\[1ex]
\begin{array}{c|c}
 & H \\
\hline
 & 1
\end{array}
\end{array}
\begin{array}{c}
% \text{one-factor with only a } V \text{ row}
% \\[1ex]
\begin{array}{c|c}
 &  \\
\hline
V & 1
\end{array}
\end{array}
\]
The first statement corresponds to the hypothesis that an association
line will be removed whenever it connects to a particular structural
element, such as a high tone or a vowel. In logical form, these
deassociation conditions can be stated as

\[
\textsc{deassoc}(x,y) \;:=\; \alpha(x,y) \;\wedge\; H(x)
\\
\textsc{deassoc}(x,y) \;:=\; \alpha(x,y) \;\wedge\; V(y)
\]
The first hypothesis predicts
that any line involving an \(H\)-tone will be deassociated, while the
second predicts that any association involving a vowel will be removed.

These hypotheses are still too general if the data contain
\(H\)-associations or vowel associations that remain unchanged, since
they would incorrectly predict deassociation in those cases.
The learner then considers two-factor hypotheses, where the triggering
environment contains two pieces of structure.

\[
\begin{array}{c}
\text{two-factor with both an } L \text{ column and a } V \text{ row}
\\[1ex]
\begin{array}{c|c}
 & L \\
\hline
V & 1
\end{array}
\end{array}
\]

This hypothesis states that deassociation applies specifically to an
association between a vowel and an \(L\) tone.

Other two-factor hypotheses add neighboring tones or neighboring
segments:

\[
\begin{array}{cc}
\begin{array}{c}
\text{two-factor with } H,L \text{ tone columns}
\\[1ex]
\begin{array}{c|cc}
 & H & L \\
\hline
 & 0 & 1
\end{array}
\end{array}
&
\begin{array}{c}
\text{two-factor with } C,V \text{ segment rows}
\\[1ex]
\begin{array}{c|c}
 &  \\
\hline
C & 0 \\
V & 1
\end{array}
\end{array}
\end{array}
\]

\[
\begin{array}{c}
\text{two-factor with adjacent } V,V \text{ segment rows}
\\[1ex]
\begin{array}{c|c}
 &  \\
\hline
V & 1 \\
V & 0
\end{array}
\end{array}
\]

If these two-factor hypotheses are still too general, the learner
continues to larger factors. For example, a three-factor hypothesis can
combine the relevant tone with the local segmental environment:

\[
\begin{array}{c}
\text{three-factor with an } L \text{ tone and adjacent } V,V \text{ rows}
\\[1ex]
\begin{array}{c|c}
 & L \\
\hline
V & 1 \\
V & 0
\end{array}
\end{array}
\]

This hypothesis states that deassociation applies to an \(L\)-tone
association on the first vowel in a \(V V\) sequence. In other words,
the learner searches for the smallest structural factor that correctly
captures all and only the observed deassociation sites.

\vfill
\bibliographystyle{apalike}
\footnotesize
\bibliography{ref}
\end{document}